### 2018-01-03 manuscript.rmd
manuscript edits through line 306, did not rebuild pdf

# Siphonophore Phylogeny

@@

Siphonophores (Fig. 1) are among the most abundant gelatinous predators in the open ocean, and have a large impact on ocean ecosystems [@pugh1984diel;@pugh1997vertical; @williams1981vertical;@purcell1981dietary;@pages2001gelatinous] [XX cite Anela and Steve's new paper]. Siphonophores, which belong to Hydrozoa (Cnidaria), are found at all depths in the ocean. The most familiar species is the Portuguese man of war *Physalia physalis*, which floats at the surface and can wash up conspicuously onto beaches [@Totton1960]. Most species are planktonic, living in the water column, where some grow to be up to 30 meters in length [@mackie1987siphonophore]. There is also a small clade of benthic siphonophores, the Rhodaliidae [@pugh1983benthic], that are tethered to the bottom for part of their lives. There are currently 188 valid described species.

Siphonophores remain poorly known, in large part because they are fragile and difficult to collect. They have, however, been of great interest for more than 150 years due to their very unique structure and development [@mapstone2014global;@mackie1987siphonophore]. Like many other cnidarians, they are colonial: they grow by incomplete asexual reproduction. Each colony arises from a single embryo that forms the protozooid, the first body. One or two growth zones then arises that asexually produces other genetically identical zooids [@carre1969etude; @carre1991complete; @carre1967; @carre1995ordre]. These zooids are each homologous to a solitary animal, but are physiologically integrated [@totton1965synopsis; @mackie1987siphonophore; @dunn2006evolution]. Siphonophores differ significantly from all other colonial animals in colony structure, development, and the high degree to which their zooids are functionally specialized [@Beklemishev1969;@cartwright2010character]. Each zooid is functionally specialized for a particular task (*e.g.*, feeding, reproducing, swimming) (Fig. 2) [@dunn2006evolution], and they are arranged in a repeating, species-specific pattern.

![Photographs of representatives of the three major groups of siphonophores. Cystonectae (A-B), Calycophorae (C-F) and Physonectae (G-K). (A) *Rhisophysa eysenhardtii*, scale bar = 1 cm. (B) *Bathyphysa conifera*, scale bar = 2cm. (C) *Hippopodius hippopus*, scale bar = 5 mm. (D) *Kephyes hiulcus*, scale bar = 2 mm. (E) *Desmophyes haematogaster*, scale bar = 5 mm. (F) *Sphaeronectes christiansonae*, scale bar = 2 mm. (G) *Nanomia bijuga*, scale bar = 1 cm. (H) *Lychnagalma utricularia*, scale bar = 1 cm. (I) *Physophora hydrostatica*, scale bar = 5 mm. (J) *Praya dubia*, scale bar = 4 cm . (K) *Apolemia* sp., scale bar = 1 cm.](figures/Figure1.jpg)

Understanding the unique ecology, morphology, and development of siphonophores requires an understanding of their phylogenetic relationships. The relationship of siphonophores to other hydrozoans has been difficult to resolve [@cartwright2010character; @Cartwright2008kl; @zapata2015phylogenomic], but there has been progress on their internal relationships. A phylogeny [@Dunn:2005dy] based on two genes (16S, 18S) from 52 siphonophore taxa advanced several long standing questions about siphonophore biology. These include the relationships of the three historically recognised groups, Cystonectae, Physonectae, and Calycophorae. The cystonects were found to be sister to all other siphonophores, while the calycophorans were nested within physonects. The name Codonophora was given to this clade of physonects and calycophorans [@Dunn:2005dy].

Major questions remained after this early work, though. There was, in particular, little support for important deep relationships within Codonophora. Resolving these relationships is key to resolving the evolution of several traits, including sexual systems (monoecy versus dioecy) and the gain and loss of particular zooids, such as palpons (Fig. 2). Here we present a broadly sampled phylogenomic analysis of Siphonophora, assessing transcriptomic data from 33 siphonophore species and 10 outgroup species (2 outgroups were subsequently excluded). Using 1,071 genes shared across species, we find strong support for several relationships within the group. Using this phylogeny, we reconstruct the evolutionary history of characters central to their unique biology, including zooid type, life history traits, and habitat.

![Schematic of the siphonophore *Nanomia bijuga*, oriented with the anterior of the colony at the top, and the ventral side to the left. Adapted from http://commons.wikimedia.org/wiki/File:Nanomia_bijuga_whole_animal_and_growth_zones.svg, drawn by Freya Goetz. (*A*) Overview of the whole mature colony. (*B*) Inset of the nectosomal growth zone with pneumatophore. A series of buds gives rise to nectophores. (*C*) Inset of the siphosomal growth zone. Probuds subdivide to give rise to zooids in repeating-units (cormidia). Cormidial boundaries are marked by a gastrozooid.](figures/Figure2.pdf)

@@

Specimens were collected in the eastern Pacific Ocean, Mediterranean, and the Gulf of California. Collection data on all examined specimens, a description of the tissue that was sampled from the colony, collection mode, sample processing details, mRNA extraction methods, sequencing library preparation methods and sequencing details are summarized in Supplementary Information. Monterey Bay and Gulf of California specimens were collected by remotely operated underwater vehicle (ROV) or during blue-water SCUBA dives. *Chelophyes appendiculata* and *Hippopodius hippopus* specimens were collected in the bay of Villefranche-sur-Mer, France, during a plankton trawl on 13 April 2011. Available physical vouchers have been deposited at the Museum of Comparative Zoology (Harvard University), Cambridge, MA, the Peabody Museum of Natural History (Yale University), New Haven, CT, or had been previously deposited at the United States National Museum (Smithsonian Institution), Washington, DC. Accession numbers are given in Supplementary Information. In cases where physical vouchers were unavailable we provide photographs to document species identity (see git repository).

@@

When possible, specimens were starved overnight in filtered seawater at temperatures close to ambient water temperatures at the time point of specimen collection (see supplementary document). mRNA was extracted directly from tissue using a variety of methods (see supplementary document): Magnetic mRNA Isolation Kit (NEB, #S1550S), Invitrogen Dynabeads mRNA Direct Kit (Ambion, #61011), Zymo Quick RNA MicroPrep (Zymo #R1050), or from total RNA after Trizol (Ambion, #15596026) extraction and through purification using Dynabeads mRNA Purification Kit (Ambion, #61006).  In case of anticipated very small total RNA quantities, only a single round of bead purification was performed. Extractions were performed according to the manufacturer's instruction. All samples were DNase treated (TURBO DNA-free, Invitrogen #AM1907;or on column DNase treatment with Zymo Quick RNA MicroPrep). Libraries were prepared for sequencing using the Illumina TruSeq RNA Sample Prep Kit (Illumina, #FC-122-1001, #FC-122-1002), the Illumina TruSeq Stranded Library Prep Kit (Illumina, #RS-122-2101) or the NEBNext RNA Sample Prep Master Mix Set (NEB, #E6110S). We collected long read paired end Illumina data for *de novo* transcriptome assembly. In the case of large tissue inputs, libraries were sequenced separately for each tissue, subsequently subsampled and pooled *in silico*. Libraries were sequenced on the HiSeq 2000, 2500, and 3000 sequencing platforms. All sequence data have been deposited in the NCBI sequence read archive (SRA) with accession number XXX.

@@

New data were analysed in conjunction with 13 publicly available datasets, with a total number of 43 species. Sequence assembly, annotation, homology evaluation, gene tree construction, parsing of genes trees to isolate orthologous sequences, and supermatrix construction were conducted with Agalma v. 1.00 [@Dunn:2013kw; @Guang202416].  This workflow integrates a variety of existing tools [@grabherr2011full, @altschul1990basic, @doi:10.1093/nar/30.7.1575, @stamatakis2006raxml, @sukumaran2010dendropy, @langmead2012fast, @li2011rsem, @talavera2007improvement, @li2009sequence, @katoh2013mafft] and new methods. Maximum likelihood analyses of the supermatrix were conducted with XX. Bayesian Inference (BI) analyses of the supermatrix were conducted using Phylobayes v. 1.7a-mpi [@lartillot2009phylobayes]. Sequence alignments, sampled and consensus trees, and voucher information are available in a git repository https://github.com/caseywdunn/siphonophore_phylogeny_2017.

Two outgroup species, *Atolla vanhoeffeni* and *Aegina citrea*, were removed from the final supermatrix due to low gene occupancy (gene sampling of 20.8% and 14.5% respectively in a 50% occupancy matrix with 2,203 genes). The final analyses presented here consider 33 siphonophore species and 8 outgroup species. This includes new data for 30 species. Summary statistics for expression libraries are given in Supplementary Information [XX This sentence seems out of place - delete? Or rephrase and move up.]. In the final analyses, we sampled 1,071 genes to generate a supermatrix with 60% occupancy and a length of 378,468 amino acids (gene occupancy matrix - supplementary figure 1).

ML analyses were conducted on the unpartitioned supermatrix using the WAG+$\Gamma$ model of amino acid substitution, and bootstrap values were estimated using `r length(bootstrap_trees)` replicates. BI was conducted using two different CAT models, CAT-Poisson and CAT-GTR [@doi:10.1093/molbev/msh112]. Two independent MCMC chains were run under the CAT-GTR model, and four independent MCMC chains were run under the CAT-Poisson model. The CAT-GTR and CAT-Poisson models did not converge [XX include the maxdiff in the text here], and only the results from the CAT-Poisson model are included here.  Visual inspection of the traces indicated that a burn in of `r burn_in` trees was sufficient for all CAT-Poisson runs. This left `r length( phylobayes_trees )` trees in the posterior.

We used the Swofford-Olsen-Waddell-Hillis (SOWH) test [@swofford1996phylogenetic] to evaluate two hypotheses: (i) physonects are monophyletic [@totton1965synopsis]; (ii) monoecious species are monophyletic [@Dunn:2005dy]. As the sexual system of *Rudjakovia* sp. is unclear, we carried out two tests of the monophyly of monoecy, one with *Rudjakovia* sp. included as a monoecious species, and one without. We used SOWHAT [@doi:10.1093/sysbio/syv055] dev. version 0.39 (commit fd68ef57) to carry out the SOWH tests in parallel with the default options and an initial sample size of 100 (source code can be found in the git repository). For each hypothesis we defined a topology with a single constrained node that was inconsistent with the most likely topology (Fig. 3). We used a threshold for significance of 0.05 and following the initial 100 samples, we evaluated the confidence interval around the p-value to determine if more samples were necessary.

Morphological character data used in trait mapping were obtained from the literature or direct observation of available voucher material. Depth distribution data was queried from the MBARI VARS ( http://www.mbari.org/products/research-software/video-annotation-and-reference-system-vars/, [@schlining2006mbari]) database. We used stochastic character mapping to infer the probable evolution of traits on the tree in R using the `phytools` package [@MEE3:MEE3169; @huelsenbeck2003stochastic]. Subsequent analyses were conducted in R and integrated into this manuscript with the `knitr` package. See Supplementary Information for R package version numbers.

@@

Most relationships received strong support across analysis methods (Figure 3A), with a couple localized exceptions (Figure 3B). The phylogenetic relationships recovered in this study are largely consistent with those found in a previous study based on two genes (16S and 18S ribosomal RNA) [@Dunn:2005dy]. Relationships that receive strong support in both studies include the placement of Cystonectae as sister to Codonophora (the clade that includes all other siphonophores), the placement of Apolemidae as sister to all other codonophorans, and the placement of Calycophorae within the paraphyletic "Physonectae". Multiple nodes that were not resolved in the previous two-gene analysis receive strong support in the present 1,071-gene transcriptome analyses. There is strong support for the Pyrostephidae as sister to all other non-apolemiid codonophorans. Within the clade that is sister to the Pyrostephidae, we find two main clades, the Calycophorae and a clade we here name Euphysonectae (Fig. 3). It includes the remaining non-apolemiid, non-pyrostephid "Physonectae". We define the Euphysonectae as the clade consisting of *Agalma elegans* and all taxa that are more closely related to it than to *Diphyes dispar*.

Euphysonectae consists of two reciprocally monophyletic groups we here provisionally refer to as Clade A and Clade B (Fig. 3A). Clade A is a group of siphonophores that all have an involucrum, or fold, that forms around the base of the tentilla [@totton1965synopsis]. This clade also shares the prescence of a descending mantle canal within the nectophores (Fig. S6), and is also monoecious (Fig. 5). Within Clade B there is low support for the placement of *Erenna richardi*, which is placed as sister to Clade A in some analyses, and more taxon sampling will be required to determine the relationship of species within this clade.

Within Clade A, *Physophora gilmeri* along with *Lychnagalma utricularia* (both not included in the previous phylogeny) are sister to the Agalmatidae, a clade restricted to *Agalma*, *Athorybia*,*Melophysa*, *Halistemma* and *Nanomia* [@Dunn:2005dy; @pugh2006taxonomic]. In the rDNA study, *Physophora hydrostatica* (the presumed sister species to *P. gilmeri*) was sister to the Forskaliidae with low support. The position of *Cordagalma cordiforme* (= *Cordagalma ordinatum*) [@pugh2016synopsis] was previously unresolved, while in this analysis *Cordagalma* sp. is in a clade with *Forskalia asymmetrica*, falling outside of the Agalmatidae. Placement of *Cordagalma* outside Agalmatidae is consistent with previous analyses using morphological and molecular data [@pugh2006taxonomic;@Dunn:2005dy].

Within the calycophorans, taxon sampling is shallower, however there is broad agreement with the previous analysis. Calycophorans have in the past been split into two groups, prayomorph and diphyomorph, based on morphology (after @mackie1987siphonophore). As in the previous study, the results presented here indicate that the prayomorphs are paraphyletic with respect to the diphyomorphs. *Craseoa lathetica* and *Desmophyes* sp. are sister to *Hippopodius hippopus* in this study, while in the previous study, the relationship between *Craseoa lathetica* and the clade including *Hippopodius hippopus* was unresolved.

>>>>

# Phylogenetic analyses of trait evolution in Siphonophora (Cnidaria)

@@

Siphonophores (Fig. 1 and 2) are among the most abundant gelatinous predators in the open ocean, and have a large impact on ocean ecosystems [@pugh1984diel;@pugh1997vertical; @williams1981vertical; @purcell1981dietary; @pages2001gelatinous;  @Choy20172116]. Siphonophores, which belong to Hydrozoa (Cnidaria), are found at all depths in the ocean. The most familiar species is the Portuguese man of war *Physalia physalis*, which floats at the surface and can wash up conspicuously onto beaches [@Totton1960]. Most species are planktonic, living in the water column, where some grow to be more than 30 meters in length [@mackie1987siphonophore]. There is also a small clade of benthic siphonophores, the Rhodaliidae [@pugh1983benthic], that are tethered to the bottom for part of their lives. There are currently 188 valid described siphonophore species.

Siphonophores remain poorly known, in large part because they are fragile and difficult to collect. They have, however, been of great interest for more than 150 years in part due to their unique structure and development [@mapstone2014global;@mackie1987siphonophore]. Like many other cnidarians, they are colonial: they grow by incomplete asexual reproduction. Each colony arises from a single embryo that forms the protozooid, the first body. One or two growth zones (Fig. 2) then arises that asexually produces other genetically identical zooids that remain attached [@carre1969etude; @carre1991complete; @carre1967; @carre1995ordre]. These zooids are each homologous to a solitary animal, but are physiologically integrated [@totton1965synopsis; @mackie1987siphonophore; @dunn2006evolution]. Siphonophores differ significantly from other colonial animals in colony structure, development, and the high degree to which their zooids are functionally specialized and arranged in precise, repeating, species-specific patterns [@Beklemishev1969;@cartwright2010character]. The functions that zooids are specialized for include feeding, reproducing, or swimming (Fig. 2) [@dunn2006evolution].

![Photographs of living siphonophores. They correspond to the clades shown in Figure 3 as follows: Cystonectae (A-B), Calycophorae (C-F, J), Clade A within Euphysonectae (G-I) and Apolemidae (K). (A) *Rhisophysa eysenhardtii*, scale bar = 1 cm. (B) *Bathyphysa conifera*, scale bar = 2cm. (C) *Hippopodius hippopus*, scale bar = 5 mm. (D) *Kephyes hiulcus*, scale bar = 2 mm. (E) *Desmophyes haematogaster*, scale bar = 5 mm. (F) *Sphaeronectes christiansonae*, scale bar = 2 mm. (G) *Nanomia bijuga*, scale bar = 1 cm. (H) *Lychnagalma utricularia*, scale bar = 1 cm. (I) *Physophora hydrostatica*, scale bar = 5 mm. (J) *Praya dubia*, scale bar = 4 cm . (K) *Apolemia* sp., scale bar = 1 cm.](figures/Figure1.jpg)

Understanding the unique ecology, morphology, and development of siphonophores requires a well-resolved phylogeny of the group. The relationship of siphonophores to other hydrozoans has been difficult to resolve [@cartwright2010character; @Cartwright2008kl; @zapata2015phylogenomic], but there has been progress on their internal relationships. A phylogeny [@Dunn:2005dy] based on two genes (16S, 18S) from 52 siphonophore taxa advanced several long standing questions about siphonophore biology. These include the relationships of the three historically recognised groups, Cystonectae, Physonectae, and Calycophorae. The Cystonectae was found to be sister to all other siphonophores, while the Calycophorae were nested within "Physonectae". The name Codonophora was given to this clade of "Physonectae" and "Calycophorae" [@Dunn:2005dy].

Major questions remained after this early work, though. There was, in particular, little support for important deep relationships within Codonophora. Understanding these relationships is key to resolving the evolution of several traits of importance, including sexual systems (monoecy versus dioecy) and the gain and loss of particular zooids, such as palpons (Fig. 2). Here we present a broadly sampled phylogenetic analysis of Siphonophora that considers transcriptomic data from 33 siphonophore species and 10 outgroup species (2 outgroups were subsequently excluded due to poor sampling). Using 1,071 genes, we find strong support for many relationships found in the earlier phylogeny [@Dunn:2005dy], and also provide new resolution for key relationships that were unresolved in that previous study. Using this phylogeny, we reconstruct the evolutionary history of characters central to the unique biology of siphonophores, including zooid type, life history traits, and habitat.

![Schematic of the siphonophore *Nanomia bijuga*, oriented with the anterior of the colony at the top of the page, and the ventral side to the left. Adapted from http://commons.wikimedia.org/wiki/File:Nanomia_bijuga_whole_animal_and_growth_zones.svg, drawn by Freya Goetz. (A) Overview of the whole mature colony. (B) Inset of the pneumatophore and nectosomal growth zone. A series of buds gives rise to nectophores. (C) Inset of the siphosomal growth zone. Probuds subdivide to give rise to zooids in repeating-units (cormidia). The gastrozooid is the posterior-most zooid within each cormidium.](figures/Figure2.pdf)

@@

Specimens were collected in the eastern Pacific Ocean, Mediterranean, and the Gulf of California. Collection data on all examined specimens, a description of the tissue that was sampled from the colony, collection mode, sample processing details, mRNA extraction methods, sequencing library preparation methods, and sequencing details are summarized in the file Supplementary data 1. Monterey Bay and Gulf of California specimens were collected by remotely operated underwater vehicle (ROV) or during blue-water SCUBA dives. *Chelophyes appendiculata* and *Hippopodius hippopus* specimens were collected in the bay of Villefranche-sur-Mer, France, during a plankton trawl on 13 April 2011. Available physical vouchers have been deposited at the Museum of Comparative Zoology (Harvard University), Cambridge, MA, the Peabody Museum of Natural History (Yale University), New Haven, CT, or had been previously deposited at the United States National Museum (Smithsonian Institution), Washington, DC. Accession numbers are given in Supplementary data 1. In cases where physical vouchers were unavailable we provide photographs to document species identity (see git repository).

@@

When possible, specimens were starved overnight in filtered seawater at temperatures close to ambient water temperatures at the time of specimen collection. mRNA was extracted directly from tissue using a variety of methods (Supplementary data 1): Magnetic mRNA Isolation Kit (NEB, #S1550S), Invitrogen Dynabeads mRNA Direct Kit (Ambion, #61011), Zymo Quick RNA MicroPrep (Zymo #R1050), or from total RNA after Trizol (Ambion, #15596026) extraction and through purification using Dynabeads mRNA Purification Kit (Ambion, #61006).  In case of anticipated very small total RNA quantities, only a single round of bead purification was performed. Extractions were performed according to the manufacturer's instruction. All samples were DNase treated (TURBO DNA-free, Invitrogen #AM1907;or on column DNase treatment with Zymo Quick RNA MicroPrep). Libraries were prepared for sequencing using the Illumina TruSeq RNA Sample Prep Kit (Illumina, #FC-122-1001, #FC-122-1002), the Illumina TruSeq Stranded Library Prep Kit (Illumina, #RS-122-2101) or the NEBNext RNA Sample Prep Master Mix Set (NEB, #E6110S). We collected long read paired end Illumina data for *de novo* transcriptome assembly. In the case of large tissue inputs, libraries were sequenced separately for each tissue, subsequently subsampled and pooled *in silico*. Libraries were sequenced on the HiSeq 2000, 2500, and 3000 sequencing platforms. Summary statistics for each library are given in the file Supplementary data 2. All sequence data have been deposited in the NCBI sequence read archive (SRA) with accession number XXX.

@@

New data were analysed in conjunction with 13 publicly available datasets, with a total number of 43 species. Sequence assembly, annotation, homology evaluation, gene tree construction, parsing of genes trees to isolate orthologous sequences, and supermatrix construction were conducted with Agalma v. 1.00 [@Dunn:2013kw; @Guang202416].  This workflow integrates a variety of existing tools [@grabherr2011full, @altschul1990basic, @doi:10.1093/nar/30.7.1575, @stamatakis2006raxml, @sukumaran2010dendropy, @langmead2012fast, @li2011rsem, @talavera2007improvement, @li2009sequence, @katoh2013mafft] and new methods. Maximum likelihood analyses of the supermatrix were conducted with [XX add raxml version and citation]. Bayesian Inference (BI) analyses of the supermatrix were conducted using Phylobayes v. 1.7a-mpi [@lartillot2009phylobayes]. Sequence alignments, sampled and consensus trees, and voucher information are available in the git repository.

Two outgroup species, *Atolla vanhoeffeni* and *Aegina citrea*, were removed from the final supermatrix due to low gene occupancy (gene sampling of 20.8% and 14.5% respectively in a 50% occupancy matrix with 2,203 genes). The final analyses presented here consider 33 siphonophore species and 8 outgroup species. This includes new data for 30 species. In the final analyses, we sampled 1,071 genes to generate a supermatrix with 60% occupancy and a length of 378,468 amino acids (Fig. S1).

ML analyses were conducted on the unpartitioned supermatrix using the WAG+$\Gamma$ model of amino acid substitution, and bootstrap values were estimated using `r length(bootstrap_trees)` replicates. BI was conducted using two different CAT models, CAT-Poisson and CAT-GTR [@doi:10.1093/molbev/msh112]. Two independent MCMC chains were run under the CAT-GTR model, and four independent MCMC chains were run under the CAT-Poisson model. The BI analyses did not converge (CAT-Poisson: maxdiff=1, meandiff=0.0148; CAT-GTR: maxdiff=1, meandiff=0.0127). Only the results from the CAT-Poisson model are presented here.Visual inspection of the traces indicated that a burn in of `r burn_in` trees was sufficient for all CAT-Poisson runs. This left `r length( phylobayes_trees )` trees in the posterior.

We used the Swofford-Olsen-Waddell-Hillis (SOWH) test [@swofford1996phylogenetic] to evaluate two hypotheses (Fig. 3C, S2): (i) "Physonectae" is monophyletic [@totton1965synopsis]; (ii) monoecious species are monophyletic [@Dunn:2005dy]. As the sexual system of *Rudjakovia* sp. is unclear, we carried out two tests of the monophyly of monoecy, one with *Rudjakovia* sp. included as a monoecious species, and one without (Fig. S2). We used SOWHAT [@doi:10.1093/sysbio/syv055] dev. version 0.39 (commit fd68ef57) to carry out the SOWH tests in parallel with the default options and an initial sample size of 100 (analysis code can be found in the git repository). For each hypothesis we defined a topology with a single constrained node that was inconsistent with the most likely topology (Fig. 3). We used a threshold for significance of 0.05 and following the initial 100 samples, we evaluated the confidence interval around the p-value to determine if more samples were necessary.

Morphological character data used in trait mapping were obtained from the literature or direct observation of available voucher material. Depth distribution data was queried from the MBARI VARS database ( http://www.mbari.org/products/research-software/video-annotation-and-reference-system-vars/) [@schlining2006mbari]. We used stochastic character mapping to infer the probable evolution of traits on the tree in R using the `phytools` package [@MEE3:MEE3169; @huelsenbeck2003stochastic]. Subsequent analyses were conducted in R and integrated into this manuscript with the `knitr` package. See Supplementary Information for R package version numbers.

@@

Most relationships received strong support across analysis methods (Fig. 3A), with a couple localized exceptions (Fig. 3B). The phylogenetic relationships recovered in this study are largely consistent with those found in a previous study based on two genes (16S and 18S ribosomal RNA) [@Dunn:2005dy]. Relationships that receive strong support in both studies include the placement of Cystonectae as sister to Codonophora (the clade that includes all other siphonophores), the placement of Apolemidae as sister to all other codonophorans, and the placement of Calycophorae within the paraphyletic "Physonectae". Multiple nodes that were not resolved in the previous two-gene analysis receive strong support in the present 1,071-gene transcriptome analyses. There is strong support for the Pyrostephidae as sister to all other non-apolemiid codonophorans. Within the clade that is sister to the Pyrostephidae, we find two main clades, the Calycophorae and a clade we here name Euphysonectae (Fig. 3A). It includes the remaining non-apolemiid, non-pyrostephid "Physonectae". We define the Euphysonectae as the clade consisting of *Agalma elegans* and all taxa that are more closely related to it than to *Diphyes dispar*.

Euphysonectae consists of two reciprocally monophyletic groups we here provisionally refer to as Clade A and Clade B (Fig. 3A). Clade A is a group of siphonophores that all have an involucrum, or fold, that forms around the base of the tentilla [@totton1965synopsis]. This clade also shares the prescence of a descending mantle canal within the nectophores (Fig. S6), and is also monoecious (Fig. 5). Within Clade B there is low support for the placement of *Erenna richardi*, which is placed as sister to Clade A in some analyses (Fig. 3B). More taxon sampling will be required to determine the relationship of species within this clade.

Within Clade A, *Physophora gilmeri* along with *Lychnagalma utricularia* (both not included in the previous phylogeny) are sister to the Agalmatidae, a clade restricted to *Agalma*, *Athorybia*,*Melophysa*, *Halistemma* and *Nanomia* [@Dunn:2005dy; @pugh2006taxonomic]. In the rDNA study, *Physophora hydrostatica* (the presumed sister species to *P. gilmeri*) was sister to the Forskaliidae with low support. The position of *Cordagalma cordiforme* (= *Cordagalma ordinatum*) [@pugh2016synopsis] was previously unresolved, while in this analysis *Cordagalma* sp. is in a clade with *Forskalia asymmetrica*, falling outside of the Agalmatidae. Placement of *Cordagalma* outside Agalmatidae is consistent with previous analyses of morphological and molecular data [@pugh2006taxonomic; @Dunn:2005dy].

Within Calycophorae, taxon sampling is shallower here than in the previous study. The calycophoran relationships that can be investigated, however, are in broad agreement with the previous analysis. Calycophorans have in the past been split into two groups, prayomorphs and diphyomorphs, based on morphology (after @mackie1987siphonophore). As in the previous study, the results presented here indicate that the prayomorphs are paraphyletic with respect to the diphyomorphs. *Craseoa lathetica* and *Desmophyes* sp. are sister to *Hippopodius hippopus* in this study, while in the previous study, the relationship between *Craseoa lathetica* and the clade including *Hippopodius hippopus* was unresolved.
